---
title: "Heart failure treatment in the last years of life"
author: "Daniel Lindholm"
output:
  html_document:
    number_sections: TRUE
    toc: TRUE
    toc_float: TRUE
---
```{r, include = FALSE}
# Setup ========================================================================
  # Default chunk options
    knitr::opts_chunk$set(
      echo = FALSE,
      message = FALSE,
      warning = FALSE,
      dev = "svg",
      fig.topcaption = TRUE,
      fig.align = "center")

  options(scipen=999)

  # Load packages
  pkgs <- c("data.table", "Epi", "fst", "ggplot2", "gt", "gtsummary",
                "labelled", "mgcv", "parallel", "popEpi", "stringr")
  vapply(pkgs, library, logical(1), character.only = TRUE, logical.return = TRUE,
       quietly = TRUE)

  # Settings for parallel processing
  setDTthreads(0L) # Use all threads with data.table
  cores <- detectCores()-1L

  # Set color scheme
  col_blrd <- c("#0072B5FF", "#BC3C29FF") 
  cols     <- c("#E18727FF", "#20854EFF", 
                "#7876B1FF", "#6F99ADFF",
                "#EE4C97FF", "#FFDC91FF")

```

```{r}
# Load data ====================================================================
  # Lists of ATC-codes, ICD codes for defining comorbidities,
  # and variable labels  
  atc        <- readRDS("/lastmeds/data/atc.rds")
  comorb_icd <- readRDS("/lastmeds/data/comorb_icd.rds")
  vars       <- readRDS("/lastmeds/data/vars.rds")

  # Main dataset (d) and comorbidities (comorb)
  d          <- read_fst("/lastmeds/data/meds.fst",   as.data.table = TRUE)
  comorb     <- read_fst("/lastmeds/data/comorb.fst", as.data.table = TRUE) 

  # Causes of death
  cod        <- read_fst("/lastmeds/data/death_icd.fst", as.data.table = TRUE) 

  # ICD10
  icd        <- fread("/lastmeds/data/icd10_2010.txt")

  # Apply labels (inefficient, since this function copies the dataset, but still)
  var_label(d)      <- vars[names(vars) %in% names(d)]
  var_label(comorb) <- vars[names(vars) %in% names(comorb)]
```
# Background
The purpose of these analyses is to investigate medication treatment patterns in
the last years before death in patients with heart failure. By looking back in 
time from the date of death, employing a Lexis-based method, to avoid 
conditioning on the future.

## Data sources
The analysis is based on Swedish nationwide registers relating to healthcare, 
from the _Swedish National Board of Health and Welfare (Socialstyrelsen)_:

- National patient register
- National cause of death register
- National prescribed drug register

## Study population
The study population is definied by:

1. **Heart failure diagnosis** at any time before death, defined as an ICD10 
code starting with **I50**, in either in-patient or specialized out-patient 
care, in any diagnosis position.
2. **Death**, with a death date on or after 2007-07-01, and an age at death of 
18 or greater.

Using _only_ I50 is based on previous research where a high diagnostic validity 
was found in the Swedish setting, when using I50 in the primary or any of the 
secondary diagnosis positions
([Schaufelberger et al 2020](https://pubmed.ncbi.nlm.nih.gov/31869528/)).

The cut-off regarding death dates was selected in order to have sufficient look 
back time (2 years or more), as the _Prescribed drugs register_ started in July 
of 2005.

## Medications
Medications under study are identified in the _Prescribed drugs register_ using 
ATC codes:

```{r}
# [TBL]: Meds definitions ======================================================  
  data.table(`Treatment class` = names(atc),
             ATC = sapply(atc, paste, collapse = "|")) |>
      gt() |>
        tab_header(title = "ATC codes for heart failure medications",
          subtitle = md("^ indicates _begins with_, | indicates _OR_"))
```

## Comorbidites
Relevant comorbidities are extracted from both _Causes of death_ and 
_National Patient registers_, using the following ICD10 codes in any position:

```{r}
# [TBL]: Comorbidities definitions =============================================
  data.table(Comorbidity = names(comorb_icd),
             ICD10 = sapply(comorb_icd, paste, collapse = "|"))[order(Comorbidity)] |>
      gt() |>
        tab_header(title = "ICD10 codes for defining comorbidities",
          subtitle = md("^ indicates _begins with_, | indicates _OR_"))
```

## Missing data
Exact death dates were missing in a very small proportion of patients (<1%).
In case of such occurrences, the following principles were employed:

- If only year is known, the date will imputed to the 1st of July of that year 
(approximately the mid point of the year)
- If only year and month is known, the date will be imputed to the 15th day of 
that month (approximately the mid point of a month).

No other imputation was performed, because missingness was very low and
considered to be negligble.


```{r}
# Derive variables =============================================================
  # [dth_yr]: year of death (rather than date) ------------------------
  d[, dth_yr := floor(dodth)]
  comorb[, dth_yr := floor(dodth)]

  # [txclass5]: collapse ACE, ARB, ARNI => 5 medication categories ----
  d[, txclass5 := ifelse(txclass %in% c("ace", "arb", "arni"),
                         "ACE/ARB/ARNI", txclass)]

  # [age_dx]: age at HF diagnosis -------------------------------------
  comorb[, age_dx := dohf - dob]
  var_label(comorb$age_dx) <- "Age at heart failure diagnosis"

  # [age_dth]: age at death -------------------------------------------
  comorb[, age_dth := dodth - dob]
  var_label(comorb$age_dth) <- "Age at death"
  d[, age_dth := dodth - dob]

  # [years_dx_dth]: years from hf diagnosis to death ------------------
  comorb[, years_dx_dth := dodth - dohf]

  # [doex]: date of exit ----------------------------------------------
  # This is the end date of the follow back time, where, in case no
  # treatment discontinuation occurred, patients will be censored.
  # Takes the maximum of these:
  #   1. date of HF diagnosis (dohf)
  #   2. 1 July 2005 (apx prescribed drugs registry start date)
  d[, doex := ifelse(dohf > cal.yr("2005-07-01"), dohf, cal.yr("2005-07-01"))]

# Select population ============================================================
  # Dataset already contains only patients who had a HF diagnosis at some point
  # before death (in any diagnosis position). 
  #
  # Prescribed medications register started in July 2005, so in order to have 
  # proper look back time (2 years or more), we will restrict the study
  # population to patients who died on 1 July 2007 or later
  #
  # Also, select patients in whom age at death equals to or exceeds 18

  d      <-      d[dodth >= cal.yr("2007-07-01") & 
                   dodth <= cal.yr("2020-12-31") &
                   age_dth >= 18 & dohf < dodth]
  comorb <- comorb[dodth >= cal.yr("2007-07-01") & 
                   dodth <= cal.yr("2020-12-31") &
                   age_dth >= 18 & dohf < dodth]

  # [FUNCTION]: cyd (cal.yr to date conversion) ----
  cyd <- function(x) structure(round((x - 1970) * 365.25), class = "Date")
```

# Population characteristics
There were `r nrow(comorb)` patients who met the criteria above, i.e. died with 
heart failure between `r cyd(min(comorb$dodth))` and `r cyd(max(comorb$dodth))`,
with a total follow-up time of `r round(d[, .SD[1L] , by = id_pat][, sum(dodth-doex)])`
person-years.

## By calendar year of death

```{r}
# [TBL]: Population characteristics (by death year) ----------------------------
  # Arrange by frequency
  tmp <- t(comorb[, 6:21][, lapply(.SD, sum)])
  tmp <- data.table(variable = rownames(tmp),
                    n = as.numeric(tmp),
                    var_number = 6:21)
  setorder(tmp, -n)
  # Create table
  comorb[, .SD, .SDcols = c(2, 22:24, tmp$var_number)] |>
    tbl_summary(by = "dth_yr") |>
      add_overall() |> 
        as_gt() |>
          tab_options(table.font.size = "65%")  
```

## By sex

```{r}
# [TBL]: Population characteristics (by sex) -----------------------------------
  comorb[, .SD, .SDcols = c(2, 22:24, tmp$var_number)] |>
    tbl_summary(by = "sex") |>
      add_overall() 
```

## Main observations

- For several of the comorbidities, there was an _increasing_ proportion over time:
  - Hypertension
  - AFib
  - Cancer
  - CKD
  - Valvular heart disease
  - Thyroid disease
  - Cardiomyopathies
  - PAH
- For __ischemic heart disease__, there was a _decreasing_ proportion over time.
- The age at diagnosis and age at death remained stable

# Causes of death

```{r}
  cod <- cod[id_pat %in% comorb$id_pat]

  # Limit to third position of ICD code
  cod[, code := substr(code, 1, 3)]

  # Left join with ICD10
  tmp <- merge(cod, icd, by = "code", all.x = TRUE)
  tmp <- tmp[, .(id_pat, code, code_text)]

  # Manually add COVID-19 text for code U07 (ICD file is from 2010)
  tmp[code == "U07", code_text := "COVID-19"]

  # For 2021, COD had not been entered yet -> remove
  tmp <- tmp[!is.na(code)]

  # Combine text and ICD code
  tmp[, cod := paste0(code_text, " (", code, ")") ]

  # Count deaths
  tmp <- tmp[, .(.N), by = "cod"]
  setorder(tmp, -N)
  
  # Create table
  tmp[1:20, .(`Cause of death (ICD10 code)` = cod)] |>
    gt() |>
      tab_header(title = "Twenty most common causes of death")

  rm(tmp)
```


# Prevalent treatment at the time of death
The proportion of patients on each medication at the time of death by calendar
year of death was estimated using a binomial model, entering year of death as
a penalized spline:
```{r}
  # [FUNCTION]: get_tx_at_death ----------------------------------------------
  # Calculates the proportion of patients on each medication at the time
  # of death, for the individuals specified in 'ids'. Default is full
  # population.        

    get_tx_at_death <- function(ids = unique(d$id_pat))
      {
        # Summarize number of deaths by year
        deaths_per_year <- d[id_pat %in% ids,
                             .SD[1L], by = .(id_pat, dth_yr)][
                                  , .N, by = .(dth_yr)][
                                      order(dth_yr)]

        # Summarize number of patients treated at time of death per class and year
        tx_per_year <- d[!is.na(txclass5) & dotxend >= dodth & id_pat %in% ids,
                         .SD[1L], by = .(id_pat, txclass5)][
                            , .(X = .N), by = .(dth_yr, txclass5)][
                                order(dth_yr)]
        
        # Join to one table
        tx_per_year[deaths_per_year, on = c("dth_yr"), N := i.N]

        tx_per_year
      }

  # [FUNCTION]: model_tx_at_death --------------------------------------------
  # Models treatment at the time of death in a binomial model entering year
  # of death as a penalized spline.     
    model_tx_at_death <- function(treatment, dataset = tx_at_death, k_val = 6){   
      gam(cbind(X, N-X) ~ s(dth_yr, k = k_val),
          family = binomial,
          data = dataset[txclass5 == treatment])
      }

  # [FUNCTION]: predict_tx_at_death ------------------------------------------
  # Makes predictions from the binomial models    
    predict_tx_at_death <- function(model, treatment, dataset = tx_at_death)
      {
          minyear <- min(dataset[txclass5 == treatment]$dth_yr)
          maxyear <- max(dataset[txclass5 == treatment]$dth_yr)
          nd <- data.frame(dth_yr = seq(minyear, maxyear, 0.2))
          pred <- as.data.table(ci.pred(model, nd) * 100)
          names(pred) <- c("est", "lci", "uci")
          pred <- cbind(pred, dth_yr = nd$dth_yr)
          pred[, txclass5 := treatment]
          pred
      }
```
```{r, fig.width = 5, fig.height = 4.5}
# [FIG]: Prevalent treatment at death by year of death -----------
  tx_at_death   <- get_tx_at_death()
  tx            <- c("ACE/ARB/ARNI", "bb", "loop", "mra", "sglt2")
  dth_models    <- lapply(tx, model_tx_at_death)
  names(dth_models) <- tx
  dth_preds     <- Map(predict_tx_at_death,
                         model = dth_models,
                         treatment = names(dth_models))
  dth_plotdata  <- rbindlist(dth_preds)
  dth_plotdata[, txclass5_fct := factor(txclass5, levels = tx,
                                  labels = c("ACEi/ARB/ARNi",
                                             "Beta blocker",
                                             "Loop diuretic", 
                                             "MRA",
                                             "SGLT2i"))]
  ggplot(dth_plotdata, aes(x = dth_yr, y = est,
                          color = txclass5_fct, fill = txclass5_fct))+
    geom_ribbon(show.legend = FALSE, 
                mapping = aes(ymin = lci, ymax = uci),
                alpha = 0.5,
                lty = 0)+
    geom_line()+
    scale_color_manual(values = cols[1:5])+
    scale_fill_manual(values = cols[1:5])+
    labs(y = "Proportion on medication (%)",
         x = "Year of death",
         color = "") +
    scale_y_continuous(expand=c(0,0), breaks = seq(0, 80, 20), limits = c(0,80))+
    scale_x_continuous(expand=c(0,0), breaks = seq(2008, 2020, 2))+
    theme_bw()+
    theme(legend.position = "top",
          legend.text = element_text(size = 6),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          aspect.ratio = 1)
```

Alternative formatting:

```{r}
dth_plotdata[, fct := "Proportion on treatment at time of death"]
death_plot <-  ggplot(dth_plotdata, aes(x = dth_yr, y = est,
                          color = txclass5_fct, fill = txclass5_fct))+
    geom_ribbon(show.legend = FALSE, 
                mapping = aes(ymin = lci, ymax = uci),
                alpha = 0.5,
                lty = 0)+
    geom_line()+
    scale_color_manual(values = cols[1:5])+
    scale_fill_manual(values = cols[1:5])+
    labs(y = "Proportion on medication (%)",
         x = "Year of death",
         color = "") +
    scale_y_continuous(expand=c(0,0), breaks = seq(0, 80, 20), limits = c(0,80))+
    scale_x_continuous(expand=c(0,0), breaks = seq(2008, 2020, 2))+
    theme_bw()+
    theme(legend.position = "right",
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          aspect.ratio = 1)+
    facet_wrap(~fct)
pdf("/lastmeds/dp2.pdf", width = 6, height = 6)
  death_plot
dev.off()

death_plot
dth_plotdata[, fct := NULL]
```

## Main observations:

- There was a substantial increase in the proportion of patients who were on
beta blocker at the time of death
- Similarly, there was a marked decrease in the proportion on a _standing 
prescription_ of loop diuretics (NB: _as needed_ prescriptions are not included)
- RAASi and MRAs were fairly stable over time
- The proportion of patients on SGLT2i at time of death was increasing in the
last few years of observation, at low numbers relative the other treatments.

# Patients on treatment relative death by year of death
The discontinuation rate was modeled using a Poisson model, and combined with
the binomial model to estimate the proportion of patients on each treatment (y axis)
by time before death (x axis) and by calendar year of death (panels).

```{r, cache = TRUE, message = FALSE, warnings = FALSE, results = FALSE}
# Treatment discontinuation --------------------------------------------------
  # [FUNCTION]: prep_data ------------------------------------------
    # Will select individuals who are not on the specified treatment
    # at the time of death. Selects the last treatment discontinuation
    # for those who had treatment before death (but not at death), and
    # one row for each patient who did not have the treatment at all.
    #
    # Then, a Lexis object is created (Lexis), and follow-back time
    # split by month (splitMulti).
    #
    # This split Lexis object will be returned.
    prep_data <- function(treatment, yr){
      tmp <- d[dth_yr == yr]

      # Select population not on treatment at the date of death
      treated_ids <- tmp[txclass5 == treatment & 
                         dotxend  >= dodth &
                         as_needed == FALSE,
                         unique(id_pat)]
      tmp <- tmp[!(id_pat %in% treated_ids)]

      # Select population to keep latest discontinuation (in case of such); 1 row,
      # or 1 row for each never on treatment
      tmp[, treated := ifelse(txclass5 != treatment | is.na(txclass5), FALSE, TRUE)]
      tmp[treated == FALSE, c("dotxstart", "dotxend") := NA]
      tmp[, c("txclass", "atc", "as_needed", "txclass5") := NULL]
      setorder(tmp, -dotxend, na.last = TRUE)
      tmp <- tmp[, .SD[1L], by = id_pat]
      
      # If treatment occured only before censoring date (e.g. before diagnosis
      # was made), set treated to FALSE
      tmp[doex > dotxend, treated := FALSE]
      
      # If failure (i.e. treatment discontinuation) occured (treated = TRUE)
      # update date of exit (doe) to take the value of last discontinuation
      # (dotxend)
      tmp[treated == TRUE, doe := dotxend]

      # Create a Lexis object
      lex <- Lexis(data        = tmp,
                   entry       = list(per = -dodth, fbt = 0),
                   exit        = list(per = -doex),
                   exit.status = factor(treated,
                                        labels = c("no tx", "tx")))

      # Split follow-back time into months
      splitlex <- splitMulti(lex, fbt = seq(0, max(lex$lex.dur), 1/12))
      splitlex
     }

  # [FUNCTION]: model_disc -----------------------------------------
    model_disc <- function(splitlex, k_val = 6){
    # Fit a Poisson model to model rate of discontinuation by follow-back time
    # modeled as a penalized spline
      bam(cbind(lex.Xst == "tx", lex.dur) ~ s(fbt, k = k_val),
          family   = poisreg,
          data     = splitlex,
          nthreads = cores,
          discrete = TRUE)
      }

  # [FUNCTION]: predict_disc ---------------------------------------
    # Make predictions
    predict_disc <- function(model, max_years){
      nd <- data.table(fbt = seq(1/12, max_years, 1/12))
      pred <- as.data.table(ci.cum(model, nd) * 100)
      names(pred) <- c("est", "lci", "uci", "se")
      pred <- cbind(pred, fbt = nd$fbt)
    }

  # [FUNCTION]: combine_models -------------------------------------
    # Combines estimates of prevalent treatment at time of death (the binomial
    # model) with the Poisson estimates of time from death to last
    # discontinuation, to estimate the proportion of patients on treatment by
    # time before death.
    combine_models <- function(treatment, yr, followback){
      # Predict discontinuation
      disc_prediction <- prep_data(treatment, yr) |>
        model_disc() |>
          predict_disc(followback)
      disc_prediction[, se := NULL]

      # Get baseline data from the binomial model predictions
      baseline <- dth_plotdata[txclass5 == treatment & dth_yr == yr]
      on_at_dth <- baseline[, est]
      
      # Scale to consider the proportion of total population
      columns <- c("est", "lci", "uci")
      disc_prediction[, (columns) := lapply(.SD,
        function(x) x * (1 - on_at_dth / 100)),
        .SDcols = columns]

      # Combine with the binomial model estimate (i.e. offset curve by
      # the proportion on medication at death)
      disc_prediction[, (columns) := lapply(.SD,
        function(x) x + on_at_dth),
        .SDcols = columns]

      results <- rbindlist(
        list(data.table(est = on_at_dth, lci = on_at_dth, uci = on_at_dth, fbt = 0),
             disc_prediction)
        )
      results[, `:=` (txclass5 = treatment, dth_yr = yr)]
      results
      }

  # Make predictions for all classes except SGLT2i
  preds <- lapply(tx[-5], function(x) {
      Map(combine_models, treatment = x, yr = 2008:2020, followback = 5)})
  
  # Make predictions for SGLT2i (special case, since it was not available during
  # the whole study period)
  pred_sglt2 <- Map(combine_models, treatment = "sglt2",
                    yr = 2015:2020, followback = 5)
  pred_sglt2 <- rbindlist(pred_sglt2)
  pred_sglt2 <- pred_sglt2[(dth_yr - fbt) > 2014]

  # Combine and tidy up
  disc_plotdata <- rbindlist(lapply(preds, rbindlist))
  disc_plotdata <- disc_plotdata[(dth_yr - fbt) > 2007]
  disc_plotdata <- rbind(disc_plotdata, pred_sglt2)
  disc_plotdata[, txclass5_fct := factor(txclass5, levels = tx,
                                      labels = c("ACEi/ARB/ARNi",
                                                 "Beta blocker",
                                                 "Loop diuretic", 
                                                 "MRA",
                                                 "SGLT2i"))]
```

```{r, fig.width = 9, fig.height = 9}
# Create plot
ggplot(disc_plotdata, aes(x = -fbt, y = est,
                          color = txclass5_fct, fill = txclass5_fct))+
  geom_ribbon(show.legend = FALSE, 
              mapping     = aes(ymin = lci, ymax = uci),
              alpha       = 0.5,
              lty         = 0)+
  geom_line()+
  scale_color_manual(values = cols[1:5])+
  scale_fill_manual(values  = cols[1:5])+
  labs(y = "Proportion on medication (%)",
       x = "Years relative death",
       color = "")+
  geom_vline(xintercept = 0, lty = 3, linewidth = 0.25)+
  scale_y_continuous(expand=c(0,0), limits = c(0, 100), breaks = seq(0, 100, 20))+
  theme_bw()+
  theme(legend.position = "top",
        panel.grid.major = element_line(size = 0.25),
        panel.grid.minor = element_blank(),
        aspect.ratio = 1) +
  facet_wrap(~ dth_yr, ncol = 5)

```

Beta blocker seem to increase over time. Plot to zoom in on beta blockers, looking 
at all years gathered into one plot. Darker line indicates more recent
calendar year.

```{r}
disc_bb <- disc_plotdata[txclass5 == "bb"]

ggplot(disc_bb, aes(x = -fbt, y = est, fill = factor(dth_yr),color = factor(dth_yr))) +
  geom_ribbon(aes(ymin = lci, ymax = uci), alpha = 0.2, lty = 0)+
  geom_line()+
  geom_text(data = disc_bb[fbt == 0],
            aes(x = 0.2, y = est, label = dth_yr),
            color = "black",
            size = 1.8)+
  geom_vline(xintercept = 0, lty = 3, linewidth = 0.25)+
  scale_fill_grey(start = 0.8, end = 0.2)+
  scale_color_grey(start = 0.8, end = 0.2)+
  scale_y_continuous(breaks = seq(50, 80, 5))+
  labs(y = "Proportion on beta blocker (%)",
       x = "Years relative death")+
  theme_bw()+
  theme(legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        aspect.ratio = 1)
```

Same kind of plot for all drug classes:

```{r}
# Define color scales
col_orange  <- colorRampPalette(c("#FFE9D1", "#E18727"), space = "Lab")
col_green   <- colorRampPalette(c("#C5E6D4", "#20854E"), space = "Lab")
col_purple  <- colorRampPalette(c("#DFDEFF", "#7876B1"), space = "Lab")
col_blue    <- colorRampPalette(c("#CFECFA", "#6F99AD"), space = "Lab")
col_magenta <- colorRampPalette(c("#FFE3F0", "#EE4C97"), space = "Lab")
disc_plotdata[, plot_colorcat := paste(txclass5, dth_yr)]
shades <- c(col_orange(13), col_green(13), col_purple(13), col_blue(13),
            col_magenta(6))


disc_plot <- ggplot(disc_plotdata, aes(x = -fbt, y = est, group = factor(dth_yr),
                          fill  = factor(plot_colorcat),
                          color = factor(plot_colorcat))) +
  geom_ribbon(aes(ymin = lci, ymax = uci), alpha = 0.2, lty = 0)+
  geom_line()+
  geom_vline(xintercept = 0, lty = 3, linewidth = 0.25)+
  scale_fill_manual(values = shades)+
  scale_color_manual(values = shades)+
  scale_x_continuous(expand = expansion(mult = c(0,0.05)),
    limits = c(-5, 0), breaks = -5:0, labels = -5:0)+
  facet_wrap(~ txclass5_fct, scales = "free_y")+
  labs(y = "Proportion on medication (%)",
       x = "Years relative death")+
  theme_bw()+
  theme(legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        aspect.ratio = 1)
pdf("/lastmeds/mp.pdf")
  disc_plot
dev.off()

disc_plot
```


## Main observations

- The discontinuation rate of _beta blockers_ decreased over time. For most
calendar years of death (where full follow-back of 5 years was available), the 
proportion of patients on beta blocker 5 years before death was estimated at 
about 80%, but the downward slope gradually diminishes over time, with later
years showing a higher proportion on beta blockers at the time of death, 
despite having a similar "start value" 5 years before.
at the time of death. 
- The discontinuation rate of loop diuretics was stable over time, but the proportion
with the medication prescribed (as a _standing prescription_) tended to decrease
- MRA treatment did not change over time
- RAAS inhibition demonstrated a higher discontinuation rate (steeper downward slope)
than other prescribed drugs, with a similar pattern observed for all calendar years

# Sensitivity analyses

## Prevalent treatment at the time of death - patients with HF diagnosis >= 5 year before death 

```{r}
# [FIG]: Prevalent treatment at death by year of death -----------
  tx_at_death   <- get_tx_at_death(comorb[years_dx_dth >= 5, id_pat])
  tx            <- c("ACE/ARB/ARNI", "bb", "loop", "mra", "sglt2")
  dth_models    <- lapply(tx, model_tx_at_death)
  names(dth_models) <- tx
  dth_preds     <- Map(predict_tx_at_death,
                         model = dth_models,
                         treatment = names(dth_models))
  dth_plotdata  <- rbindlist(dth_preds)
  dth_plotdata[, txclass5_fct := factor(txclass5, levels = tx,
                                  labels = c("ACEi/ARB/ARNi",
                                             "Beta blocker",
                                             "Loop diuretic", 
                                             "MRA",
                                             "SGLT2i"))]
  ggplot(dth_plotdata, aes(x = dth_yr, y = est,
                          color = txclass5_fct, fill = txclass5_fct))+
    geom_ribbon(show.legend = FALSE, 
                mapping = aes(ymin = lci, ymax = uci),
                alpha = 0.5,
                lty = 0)+
    geom_line()+
    scale_color_manual(values = cols[1:5])+
    scale_fill_manual(values = cols[1:5])+
    labs(y = "Proportion on medication (%)",
         x = "Year of death",
         color = "") +
    scale_y_continuous(expand=c(0,0), breaks = seq(0, 80, 20), limits = c(0,80))+
    scale_x_continuous(expand=c(0,0), breaks = seq(2008, 2020, 2))+
    theme_bw()+
    theme(legend.position = "top",
          legend.text = element_text(size = 6),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          aspect.ratio = 1)
```

## Prevalent treatment at the time of death - patients with ischemic heart disease 

```{r}
# [FIG]: Prevalent treatment at death by year of death -----------
  tx_at_death   <- get_tx_at_death(comorb[ihd == TRUE, id_pat])
  tx            <- c("ACE/ARB/ARNI", "bb", "loop", "mra", "sglt2")
  dth_models    <- lapply(tx, model_tx_at_death)
  names(dth_models) <- tx
  dth_preds     <- Map(predict_tx_at_death,
                         model = dth_models,
                         treatment = names(dth_models))
  dth_plotdata  <- rbindlist(dth_preds)
  dth_plotdata[, txclass5_fct := factor(txclass5, levels = tx,
                                  labels = c("ACEi/ARB/ARNi",
                                             "Beta blocker",
                                             "Loop diuretic", 
                                             "MRA",
                                             "SGLT2i"))]
  ggplot(dth_plotdata, aes(x = dth_yr, y = est,
                          color = txclass5_fct, fill = txclass5_fct))+
    geom_ribbon(show.legend = FALSE, 
                mapping = aes(ymin = lci, ymax = uci),
                alpha = 0.5,
                lty = 0)+
    geom_line()+
    scale_color_manual(values = cols[1:5])+
    scale_fill_manual(values = cols[1:5])+
    labs(y = "Proportion on medication (%)",
         x = "Year of death",
         color = "") +
    scale_y_continuous(expand=c(0,0), breaks = seq(0, 80, 20), limits = c(0,80))+
    scale_x_continuous(expand=c(0,0), breaks = seq(2008, 2020, 2))+
    theme_bw()+
    theme(legend.position = "top",
          legend.text = element_text(size = 6),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          aspect.ratio = 1)
```

# Reproducibility
All calculations were performed at Bianca, a part of the UPPMAX high-performance 
computing infrastructure. To ensure reproducibility, the analysis environment was containerized,
such that operating system, R version, and R package versions
were locked; using _Apptainer_ (formerly known as Singularity). For details see session info below:
```{r}
sessionInfo()
```